Chasing \rainbows{} with the Habitable Worlds Observatory

### 2023-04-17 src/tex/ms.tex
[showyourwork] overleaf sync

\subsection{\hwo{} and Stellar Sample}
%\textcolor{blue}{Merge the 2 sections here into "Assumed technical specifications and sample"? I don't think this needs to be 2 subsections}
\hwo{} is still in the early stages of development, and the exact telescope and instrument design is as of yet undecided. 
%
In this work, we assume a \SI{6}{\meter} primary mirror, which is the currently favoured design for the observatory.
%
The observatory will utilise a coronagraph to suppress the stellar light and facilitate high-contrast observations of faint accompanying exoplanet. 
%
The \IWA{} of the coronagraph is currently undecided and will be driven by both the science requirements and technological limitations. 
%In this work, we investigate the scattering phase coverage obtainable for several choices for the \IWA{} of the observatory. 
%We also consider the Outer Working Angle (OWA) but for most targets this is not a limiting factor. 
%From the HabWorlds report. Inner Working Angle: The IWA defines the region near the star that cannot be accessed for direct imaging due to a coronagraphic mask, starshade obscuration, or an interferometric null. It depends on the architecture of the telescope, the starlight suppression system used, and the observation wavelength. While a nominal value could be defined by making assumptions about the design for Habitable Worlds Observatory, we choose instead to derive the IWA from the star list itself and Astro2020’s requirement that ~100 cumulative habitable zones be surveyed by direct imaging. As shown in the main table, an IWA near ~70 mas will be needed to access ~100 cumulative habitable zones, and would need to be achieved at all wavelengths of interest for spectral characterization. For any specific telescope and starlight suppression system architecture the IWA is typically proportional to wavelength; thus, the number of accessible HZs will strongly decrease as IWA(lambda) increases.
For ease of comparison, \cref{tab:IWA_OWA} lists the angular separations corresponding to multiples of $\lambda / D$ for a \SI{6}{\meter} telescope at representative visible and near infrared wavelengths of $\lambda = \SI{600}{\nano\meter}$ and $\SI{1}{\micro\meter}$.

@@

In the non-eccentric case (i.e., with a circular orbit and where the coronagraphic mask centred on the star), minimum ($\varphi_\mathrm{min}$) and maximum ($\varphi_\mathrm{max}$) scattering phase angles accessible are symmetric about 90 degrees. 
%
We define the angle $\Delta \varphi$ such that: 
\begin{equation}
 \label{eq:Delta_phi}
    \Delta \varphi 
    = \varphi_\mathrm{max} - \varphi_\mathrm{min}
    =  2(\varphi_\mathrm{max} - 90^\circ) 
    =  2(90^\circ-\varphi_\mathrm{min} ) 
\end{equation}
% \begin{equation}
%     \label{eq:Delta_phi}
%     \Delta \varphi 
%     = \varphi_\mathrm{max} - \varphi_\mathrm{min}
%     =  2(\varphi_\mathrm{max} - 90) 
%     =  2(\varphi_\mathrm{min} + 90)
% \end{equation}

% {\color{Green} Victor Trees: Do you mean:
%INDEED, THERE MUST HAVE BEEN A TYPO, THX!
% %
% \begin{equation}
%     \Delta \varphi 
%     = \varphi_\mathrm{max} - \varphi_\mathrm{min}
%     =  2(\varphi_\mathrm{max} - 90^\circ) 
%     =  2(90^\circ-\varphi_\mathrm{min} ) \notag
% \end{equation}
%
% }

@@

\begin{figure*}[t]
    \centering
    \includegraphics[width=0.95\textwidth]{figures/accessible_phase_angles.png}  
    \caption{
        \todo{shouldn't lambda/D for 600nm give 21, 41, 62, 83 mas?} 
        Cumulative distributions of the most extreme scattering phase angle accessible for different \IWA{} where the solid lines are for randomly inclined circular orbits and the dashed lines are for randomly orientated elliptical orbits.
        %
        The top x axis indicates the minimum scattering phase angle and the bottom x axis indicates the maximum.
        %
        These are symmetric about quadrature (90 degrees) \todo{is this true for elliptical orbits?}.
        %
        The y axis indicates the number of systems divided by the number of Monte Carlo samples and is thus normalised to the number of systems in the target list.
        %
        The secondary y axis indicates the number of systems assuming only 24 percent of them contain an Earth-like exoplanet in the habitable zone.
    }
    \label{fig:betaallofit}
    \script{plot_n_versus_phase.py}
\end{figure*}

@@

\subsection{Simulating contrast curves}
\todo{Kim and David: is this done?}

@@

\subsection{The effect of eccentricity}
\label{sec:result_eccentricity}
As shown in \cref{fig:betaallofit}, allowing the planet's orbit to be eccentric can have a noticeable affect on the number of systems that can be observed at each scattering phase angle.
%
This is because, Kepler's second law ensures that eccentric planets spend more time at orbital separations greater than their semi-major axis, and hence are more likely to peak out from behind the coronagraph \IWA{} and extreme orbital phases (albeit with worse contrast due to the greater star--planet separation).
%
The eccentricity does not significantly effect the number of systems for smaller inner working angles as the systems in \hwo{} target list where chosen based on their on-sky separation quadrature and therefore the shape of the orbit has little affect for \IWA{} smaller than cut used in the selection. It is, however, important to account for orbital eccentricity when the \IWA{} of the coronoagraph is larger than the \IWA{} used to filter the target list otherwise the number of systems observable at extreme phase angles would be under predicted.

@@

The Rayleigh scattering feature appears near quadrature and is therefore observable in the most systems.
%
This feature is caused by atmospheric gases and can therefore be used to determine if a planet has an atmosphere.
%
The rainbow feature appears further from quadrature but will still be observable in a significant fraction of systems.
%
This feature will only be present if the atmosphere contains spherical droplets, potentially indicating the presence of liquid water.
%
The position and shape of the peak is sensitive to the properties of the droplets allowing their nature to be determined if the peak of the rainbow feature can be reached.
%
The ocean glint feature occurs even further from quadrature and so is observable in fewer systems. In most systems, only the first half of the feature is observable but this can still be used to identify a liquid ocean on the surface of the planet.
%
A key indicator of habitability. Finally, it is possible to view the glory scattering feature in one system, Alpha Cen A.
%
\todo{maybe we want a table of some numbers in this section}

\begin{table}
    \centering
    \caption{
        Number of stars around which the peak phase angle of each phenomena is detectable, expressed in $mas$ and $\lambda/D$ for $\lambda=600$nm. 
    }
    \label{tab:nstars_detect}
    \begin{tabular}{ c c c c } 
    \toprule
      & \multicolumn{3}{c}{IWA}
    \\
     & 21\,$mas$ & 41\,$mas$ & 62\,$mas$
    \\
     & $2\lambda/D$ & $3\lambda/D$ & $4\lambda/D$ 
    \\
    \midrule
    \midrule
    Rainbows & 90 & 42 & 18 \\
    Rayleigh & 152 & 115 & 65 \\
    Ocean Glint & 43 & 14 & 5 \\
    % Glory & 0 & 5 & 10 \\
    \bottomrule
    \end{tabular}
\end{table}

@@

The accessible scattering phase angles is not the factor that needs to be considered when determining what scattering features are observable.
%
The contrast of the planet is also an important factor to consider.
%
\cref{fig:contrasts} shows approximate contrast curves for three of the stars in the \hwo{} target list assuming edge-on orbits.
%
The over all contrast varies significantly for each system and throughout the orbit.
%
The observability of these planets will depend on the star light suppression of the coronagraph which is influenced by its \IWA{}.
%
It is not possible to predict how much the state of the art of coronagraphy will have advanced by the time \hwo{} is built.
%
However, when its design is being finalised, it will be important to consider both the accessible scattering phase angles and contrast limits when deciding on the \IWA{} of the coronagraph, if these features are to be observed.

%\subsection{The glory of Alpha Cen}%Halelujah
%\label{sec:ealpha-cen}
%\textbf{Glory science, what it tells us (Kim)}
% Kim will write para on science of glory (what tells us, vaguely why appears here)

@@

%\begin{figure*}
%    \centering
%    \includegraphics[width=0.6\textwidth]{figures/features_plot_peak.png}  
%    \caption{
%        Number of systems vs inner working angle
%    }
%    \label{fig:accessible_phase_angles}
%\end{figure*}

@@

%\subsection{Contrast Curves}
%\label{sec:contrast}

@@

We primarily focused on the range of scattering phase angles and therefore which scattering phenomena could potentially be observed given the limitation of the coronagraphic \IWA{}.
%
Coronagraphy is a rapidly advancing field, and it is hard to predict what the capabilities of \hwo{} will be.
%
Assuming a \SI{60}{\mas} \IWA{}, all 160 systems in the NASA Exoplanet Exploration Program's Mission Star List for the Habitable Worlds Observatory would be observable at scattering phases with Rayleigh features, but the rainbows of water clouds would only be accessible in $\sim$55 systems, and ocean glint in $\sim$20 systems.
%
The habitability of Earth analogs in the remaining 140 systems would have to be established by less direct means like rotational mapping to identify continents and oceans \citep[e.g.,][]{2009ApJ...700..915C,lustig2019}.   
%

The number of habitable zones that can be probed by \hwo{} increases with decreasing \IWA{}. 
%
The desire to detect and characterize a large number of Earth-like exoplanets, and the intrinsic frequency of such planets, will drive the choice of coronagraphic \IWA{} for this mission.
%
We have shown that only $\sim$10\% of Earth-like exoplanets imaged with \hwo{} could be imaged at a sufficiently wide range of phase angles to establish the presence of surface liquid water via its specular reflection.  
%
The habitability of remaining targets would have to be established by less direct means.
%
While we briefly considered the contrast ratio of the scattering features, we did not investigate the feasibility of detecting them given their contrast. 
%
This will be an important task once the instrument design and capabilities are better defined and could drive the design of the coronagraph.

>>>>

\subsection{\hwo{} and Stellar Sample}
%\textcolor{blue}{Merge the 2 sections here into "Assumed technical specifications and sample"? I don't think this needs to be 2 subsections}
\hwo{} is still in the early stages of development, and the exact telescope and instrument design is as of yet undecided. 
%
In this work, we assume a \SI{6}{\meter} primary mirror, as suggested in the The National Academy of Sciences Astronomy \& Astrophysics 2020 Decadal Survey.
%
The observatory will utilise a coronagraph to suppress the stellar light and facilitate high-contrast observations of faint accompanying exoplanet. 
%
The \IWA{} of the coronagraph is currently undecided and will be driven by both the science requirements and technological limitations. 
%In this work, we investigate the scattering phase coverage obtainable for several choices for the \IWA{} of the observatory. 
%We also consider the Outer Working Angle (OWA) but for most targets this is not a limiting factor. 
%From the HabWorlds report. Inner Working Angle: The IWA defines the region near the star that cannot be accessed for direct imaging due to a coronagraphic mask, starshade obscuration, or an interferometric null. It depends on the architecture of the telescope, the starlight suppression system used, and the observation wavelength. While a nominal value could be defined by making assumptions about the design for Habitable Worlds Observatory, we choose instead to derive the IWA from the star list itself and Astro2020’s requirement that ~100 cumulative habitable zones be surveyed by direct imaging. As shown in the main table, an IWA near ~70 mas will be needed to access ~100 cumulative habitable zones, and would need to be achieved at all wavelengths of interest for spectral characterization. For any specific telescope and starlight suppression system architecture the IWA is typically proportional to wavelength; thus, the number of accessible HZs will strongly decrease as IWA(lambda) increases.
For ease of comparison, \cref{tab:IWA_OWA} lists the angular separations corresponding to multiples of $\lambda / D$ for a \SI{6}{\meter} telescope at representative visible and near infrared wavelengths of $\lambda = \SI{600}{\nano\meter}$ and $\SI{1}{\micro\meter}$.

@@

In the non-eccentric case (i.e., with a circular orbit and where the coronagraphic mask centred on the star), minimum ($\varphi_\mathrm{min}$) and maximum ($\varphi_\mathrm{max}$) scattering phase angles accessible are symmetric about 90 degrees. 
%
We define the angle $\Delta \varphi$ such that: 
\begin{equation}
 \label{eq:Delta_phi}
    \Delta \varphi 
    = \varphi_\mathrm{max} - \varphi_\mathrm{min}
    =  2(\varphi_\mathrm{max} - 90^\circ) 
    =  2(90^\circ-\varphi_\mathrm{min} ) 
\end{equation}

@@

\begin{figure*}%[t]
    \centering
    \includegraphics[width=0.95\textwidth]{figures/accessible_phase_angles.png}  
    \caption{
        \todo{shouldn't lambda/D for 600nm give 21, 41, 62, 83 mas?} 
        Cumulative distributions of the most extreme scattering phase angle accessible for different \IWA{} where the solid lines are for randomly inclined circular orbits and the dashed lines are for randomly orientated elliptical orbits.
        %
        The top x axis indicates the minimum scattering phase angle and the bottom x axis indicates the maximum.
        %
        These are symmetric about quadrature (90 degrees).
        %
        The y axis indicates the number of systems divided by the number of Monte Carlo samples and is thus normalised to the number of systems in the target list.
        %
        The secondary y axis indicates the number of systems assuming only 24 percent of them contain an Earth-like exoplanet in the habitable zone.
    }
    \label{fig:betaallofit}
    \script{plot_n_versus_phase.py}
\end{figure*}

@@

\subsection{Simulating contrast curves}

@@

\subsection{The effect of eccentricity}
\label{sec:result_eccentricity}
As shown in \cref{fig:betaallofit}, allowing the planet's orbit to be eccentric can have a noticeable affect on the number of systems that can be observed at each scattering phase angle.
%
This is because, Kepler's second law ensures that eccentric planets spend more time at orbital separations greater than their semi-major axis, and hence are more likely to peak out from behind the coronagraph \IWA{} and extreme orbital phases (albeit with worse contrast due to the greater star--planet separation).
%
The eccentricity does not significantly effect the number of systems for \IWA{} smaller than cut used in the selection of the systems in \hwo{} target list. It is, however, important to account for orbital eccentricity when the \IWA{} of the coronoagraph is larger than the \IWA{} used to filter the target list otherwise the number of systems observable at extreme phase angles would be under predicted.

@@

The Rayleigh scattering feature, which can used to determine if a planet has an atmosphere, appears near quadrature and is therefore observable in the most systems.
%
The rainbow feature appears further from quadrature but will still be observable in a significant fraction of systems.
%
The position and shape of the peak is sensitive to the properties of the droplets allowing their nature to be determined if the peak of the rainbow feature can be reached.
%
The ocean glint feature occurs even further from quadrature and so is observable in fewer systems. 
%
In most systems, only the first half of the feature is observable but this can still be used to identify a liquid ocean on the surface of the planet.
%
A key indicator of habitability. Finally, it is possible to view the glory scattering feature in one system, Alpha Cen A.

\begin{table}
    \centering
    \caption{
        Number of stars around which the peak phase angle of each phenomena is detectable, expressed in $mas$ and $\lambda/D$ for $\lambda=600$nm and  $D=6$m. 
    }
    \label{tab:nstars_detect}
    \begin{tabular}{ c c c c } 
    \toprule
      & \multicolumn{3}{c}{IWA}
    \\
     & 21\,$mas$ & 41\,$mas$ & 62\,$mas$
    \\
     & $2\lambda/D$ & $3\lambda/D$ & $4\lambda/D$ 
    \\
    \midrule
    \midrule
    Rainbows & 90 & 42 & 18 \\
    Rayleigh & 152 & 115 & 65 \\
    Ocean Glint & 43 & 14 & 5 \\
    % Glory & 0 & 5 & 10 \\
    \bottomrule
    \end{tabular}
\end{table}

@@

The accessible scattering phase angles is not the only factor that needs to be considered when determining what scattering features are observable.
%
The contrast of the planet is also an important factor to consider.
%
\cref{fig:contrasts} shows approximate contrast curves for three of the stars in the \hwo{} target list assuming edge-on orbits.
%
The over all contrast varies significantly for each system and throughout the orbit.
%
The observability of these planets will depend on the star light suppression of the coronagraph which is influenced by its \IWA{}.
%
It is not possible to predict how much the state of the art of coronagraphy will have advanced by the time \hwo{} is built.
%
However, when its design is being finalised, it will be important to consider both the accessible scattering phase angles and contrast limits when deciding on the \IWA{} of the coronagraph, if these features are to be observed.

@@

@@

@@

We primarily focused on the range of scattering phase angles and therefore which scattering phenomena could potentially be observed given the limitation of the coronagraphic \IWA{}.
%
Coronagraphy is a rapidly advancing field, and it is hard to predict what the capabilities of \hwo{} will be.
%
Assuming a \SI{62}{\mas} \IWA{}, all 160 systems in the NASA Exoplanet Exploration Program's Mission Star List for the Habitable Worlds Observatory would be observable at scattering phases with Rayleigh features, but the rainbows of water clouds would only be accessible in $\sim$42 systems, and ocean glint in $\sim$14 systems.
%
The habitability of Earth analogs in the remaining 146 systems would have to be established by less direct means like rotational mapping to identify continents and oceans \citep[e.g.,][]{2009ApJ...700..915C,lustig2019}.   
%
The number of habitable zones that can be probed by \hwo{} increases with decreasing \IWA{}. 
%
The desire to detect and characterize a large number of Earth-like exoplanets, and the intrinsic frequency of such planets, will drive the choice of coronagraphic \IWA{} for this mission.
%
We have shown that with a $6$m primary mirror and \IWA $= 3\lambda/D$ at $600$nm coronograph, only $\sim$10\% of Earth-like exoplanets imaged with \hwo{} could be imaged at a sufficiently wide range of phase angles to establish the presence of surface liquid water via its specular reflection.  
%
The habitability of remaining targets would have to be established by less direct means.
%
This could be improved by increasing the primary mirror diameter or decreasing the \IWA of the coronograph.
%
While we briefly considered the contrast ratio of the scattering features, we did not investigate the feasibility of detecting them given their contrast. 
%
This will be an important task once the instrument design and capabilities are better defined and could drive the design of the coronagraph and the primary mirror diameter.
